Architectures for workflow integration in science gateways

### 2016-04-12 paper.tex
Wrote section 4.4

@@

\emph{Integration effort} measures the development effort required to build
the system, assuming that none of the components could initially
communicate with each other. It is measured by counting the number of
interactions to develop and components to modify on the architectural
diagram. It breaks down to 2 metrics that relate to the direction of
interaction arrows on the architectural diagrams:
\begin{itemize}[itemsep=0cm]
\item Interactions from the science gateway. 
\item Iteractions from the workflow engine. 
\end{itemize}

@@

\emph{Modularity} measures the ability to replace elements in the
architecture, or to integrate new elements. Elements include
workflow engines, workflows and infrastructure. It breaks down in 4
metrics:
\begin{itemize}[itemsep=0cm]
\item New engine type: number of interactions or components to modify
  to integrate a new workflow engine type in the
  architecture. Modification of an interaction is deemed necessary
  when the parameters involved in this interaction are
  modified. Modification of a component is required when its code
  needs to be modified or recompiled.
\item New engine version: number of interactions or components to
  modify to integrate a new version of a workflow engine in the
  architecture, assuming that another version of the same engine type
  is already available. We assume that different versions of a
  workflow engine share the same interface, i.e. they can be invoked
  using the same software. When this is not the case, the different
  versions have to be considered as different engine types.
\item New infrastructure: number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture.
\item New workflow: number of interactions required to integrate a new
  workflow in the architecture.
\item Meta-workflow: ability to describe meta-workflows from existing
  workflows (1: not available; 0: available).
\end{itemize}

@@

\paragraph{Integration} Integrating a workflow engine as a service
requires to develop 3 interactions and 1 component. First, a workflow
service has to be developed and interfaced with the infrastructure
through interactions \texttt{b} and \texttt{c$_2$}
(\texttt{I$_2$=3}). In addtion, the science gateway has to be extended
to call the engine interface \texttt{d} (\texttt{I$_1$=1}).

\paragraph{Robustness} Workflow execution requires 3 components: the
science gateway, the workflow service, and the infrastructure
(\texttt{R$_1$=3}). It involves the 4 interactions \texttt{b},
\texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
(\texttt{R$_2$=4}). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{R$_3$=0}).

\paragraph{Modularity} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{M$_1$=4}). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{M$_2$=0}). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{M$_3$=3}). New workflows are added in the science gateway or
in the workflow engine through interaction \texttt{a} only
(\texttt{M$_4$=1}). Meta-workflow are not supported by default
(\texttt{M$_5$=1}).

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$=1}). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available (\texttt{S$_2$=1}). Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_3$=1}). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_4$=0}).

@@

Science gateway has to make sure that the machine on which a workflow
task executes is reliable as several sub-tasks will depend on the
success of this task. Or workflow engine task has to be recoverable.

Performance: there could be a deadlock when there is contention. Also,
the workflow task has to be reliable (for instance, proper walltime
estimation), which is difficult because these tasks are
long. Restarting a workflow task may mean that all the tasks in the
workflow have to be restarted, depending on the capability of the
workflow engine. This complexifies task scheduling (a system could
ignore these issues but they may backfire).

@@

\begin{table*}
\centering
\begin{tabular}{rcccccc}
\textbf{Metric}                      & \textbf{Tight}
                                     & \textbf{Service}
                                     & \textbf{Sub-task}
                                     & \textbf{Pool}
                                     & \textbf{Nested}
                                     & \textbf{Import} \\
\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration effort}}\\
Science gateway -- \texttt{I$_1$}    & \multirow{2}{*}{3}
                                     & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99FF99}2  
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}3 \\
Workflow engine -- \texttt{I$_2$}    &
                                     & \cellcolor[HTML]{99FF99}3 
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99FF99}1 \\
\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99FF99}\textbf{4}
                                     & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4} \\

@@

@@

                                     & \cellcolor[HTML]{99FF99}\textbf{16}

>>>>

\todo{Interactions c1 and c2 are merged in c since this is done
  completely by the science gateway}

@@

\todo{Refer to the metric codes.}

\emph{Integration} measures the development effort required to build
the architecture, assuming that the science gateway and infrastructure
are already integrated. It is measured by counting on the
architectural diagram the number of interactions and components to
develop or modify. It breaks down to 4 metrics that quantify the
developments related to:
\begin{itemize}[itemsep=0cm]
\item Science gateway (\texttt{I$_1$}),
\item Workflow engine (\texttt{I$_2$}),
\item Infrastructure (\texttt{I$_3$}),
\item Other components (\texttt{I$_4$}).
\end{itemize}
Interactions are counted only once, in the component from which they
originate.

@@

\emph{Modularity} measures the ability to replace elements in the
architecture, or to integrate new elements. Elements include
workflow engines, workflows and infrastructure. It breaks down in 4
metrics:
\begin{itemize}[itemsep=0cm]
\item New engine type: number of interactions or components to modify
  to integrate a new workflow engine type in the
  architecture. Modification of an interaction is deemed necessary
  when the parameters involved in this interaction are
  modified. Modification of a component is required when its code
  needs to be modified or recompiled (science gateway or workflow
  service), or when a new piece of software has to be installed
  (infrastructure only).
\item New engine version: number of interactions or components to
  modify to integrate a new version of a workflow engine in the
  architecture, assuming that another version of the same engine type
  is already available. We assume that different versions of a
  workflow engine share the same interface, i.e. they can be invoked
  using the same software. When this is not the case, the different
  versions have to be considered as different engine types.
\item New infrastructure: number of interactions or components to
  modify to integrate a new type of infrastructure in the
  architecture.
\item New workflow: number of interactions required to integrate a new
  workflow in the architecture.
\item Meta-workflow: ability to describe meta-workflows from existing
  workflows (1: not available; 0: available).
\end{itemize}

@@

\paragraph{Integration} Integrating a workflow engine as a service
requires to develop 3 interactions and 1 component. First, a workflow
service has to be developed and interfaced with the infrastructure
through interactions \texttt{b} and \texttt{c$_2$}
(\texttt{I$_2$}=3). In addtion, the science gateway has to be extended
to call the engine interface \texttt{d} (\texttt{I$_1$}=1).

\paragraph{Robustness} Workflow execution requires 3 components: the
science gateway, the workflow service, and the infrastructure
(\texttt{R$_1$}=3). It involves the 4 interactions \texttt{b},
\texttt{$c_1$}, \texttt{$c_2$} and \texttt{d}
(\texttt{R$_2$}=4). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{R$_3$}=0).

\paragraph{Modularity} Adding a new type of workflow engine requires
to implement the corresponding workflow service, to modify interaction
\texttt{d}, and to implement interactions \texttt{b} and
\texttt{$c_2$} (\texttt{M$_1$}=4). New engine versions can be added by
updating the workflow service without modifying any interaction
(\texttt{M$_2$}=0). Adding a new type of infrastructure requires
updates in interactions \texttt{b}, \texttt{$c_1$} and \texttt{$c_2$}
(\texttt{M$_3$}=3). New workflows are added in the science gateway or
in the workflow engine through interaction \texttt{a} only
(\texttt{M$_4$}=1). Meta-workflow are not supported by default
(\texttt{M$_5$}=1).

\paragraph{Scalability} The service architecture supports multiple
engine instances through multiple workflow services. In VIP for
instance, this feature has been available from release 1.17 (April
2016). A basic load-balancing mechanism is available that sends new
workflow executions to the engine instance that has the least active
executions. To avoid ``black-hole'' syndromes created by failing
engine instances, engine instances are automatically disabled when
workflows cannot be submitted to them. Adding a new engine instance,
however, requires manual intervention to declare the new instance in
the science gateway (\texttt{S$_1$}=1). Consequently, elastic engines
are difficult to implement because they require a mechanism to update
the science gateway configuration when a new engine instance is
available (\texttt{S$_2$}=1). Distributing the execution of a single
workflow in multiple engines is usually not possible unless the
workflow engine has specific abilities (\texttt{S$_3$}=1). The
scheduling of tasks on the infrastructure is as complex as in any
other architecture since the workflow engine might implement any kind
of scheduling policy (\texttt{S$_4$}=0).

@@

\paragraph{Integration} Integrating a workflow engine as a sub-task
only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
and to install the workflow engine on the infrastructure
(\texttt{I$_3$}=1). A new task type is created in the science gateway
only when a new workflow has to be integrated (see M$_4$), but this is
not required to integrate the engine itself (\texttt{I$_1$}=0).

\paragraph{Robustness} Only 2 components and 3 interactions are
involved in workflow execution (\texttt{R$_1$}=2,
\texttt{R$_2$}=3). Obtaining fine-grained information about workflow
activities is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{R$_3$}=1).

\paragraph{Modularity} Integrating a new type of workflow engine
requires to develop interaction \texttt{e} and to install the engine
on the infrastructure (\texttt{M$_1$}=2). Updating an engine version
requires updates only on the infrastructure (\texttt{M$_2$}=1). Adding
a new infrastructure requires to update interactions \texttt{b} and
\texttt{c} in the science gateway (\texttt{M$_3$}=2). New workflows are integrated by
creating a new task in the science gateway through interaction
\texttt{a} (\texttt{M$_4$}=1), and meta-workflows are not available
(\texttt{M$_5$}=1).

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0,
\texttt{S$_2$}=0). Such scalability properties are one of the major
interests of the sub-task architecture. Distributed engines are not
supported by default (\texttt{S$_3$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_4$}=1). Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for sub-tasks which may generate issues such as
selection of wrong batch queues on clusters. Finally, the
interdependencies between the workflow task and its sub-tasks may
create deadlocks when there is contention. Say for instance that only
1 computing resource is available for the science gateway and that the
workflow task is running on it and submits sub-tasks, then the
sub-tasks could only execute when the resource is available, which
will never happen because the workflow task will not complete until
the sub-tasks complete. This configuration can be generalized to an
infrastructure with $n$ resources where $n$ workflows are
submitted. In practice, however, the number of submitted workflows
usually remains lower than the number of computing resources available
on this infrastructure, which makes such deadlocks unlikely to happen.

@@

\begin{table*}
\centering
\begin{tabular}{rcccccc}
\textbf{Metric}                      & \textbf{Tight}
                                     & \textbf{Service}
                                     & \textbf{Sub-task}
                                     & \textbf{Pool}
                                     & \textbf{Nested}
                                     & \textbf{Import} \\
\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Integration}}\\
Science gateway -- \texttt{I$_1$}    & \multirow{2}{*}{3}
                                     & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99FF99}0  
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99FF99}2
                                     & \cellcolor[HTML]{99AA99}3 \\
Workflow engine -- \texttt{I$_2$}    &
                                     & \cellcolor[HTML]{99FF99}3 
                                     & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99FF99}1 \\
Infrastructure -- \texttt{I$_3$}   & \cellcolor[HTML]{99FF99}0
                                     & \cellcolor[HTML]{99FF99}0
                                     & \cellcolor[HTML]{99FF99}1
                                     & \cellcolor[HTML]{99AA99}0
                                     & \cellcolor[HTML]{99AA99}0
                                     & \cellcolor[HTML]{99FF99}0 \\
Other -- \texttt{I$_4$}              & \cellcolor[HTML]{99FF99}0
                                     & \cellcolor[HTML]{99FF99}0
                                     & \cellcolor[HTML]{99FF99}0
                                     & \cellcolor[HTML]{99AA99}2
                                     & \cellcolor[HTML]{99AA99}0
                                     & \cellcolor[HTML]{99FF99}0 \\
\textbf{Total}                       & \cellcolor[HTML]{99FF99}\textbf{3}
                                     & \cellcolor[HTML]{99FF99}\textbf{4}
                                     & \cellcolor[HTML]{99FF99}\textbf{2}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4}
                                     & \cellcolor[HTML]{99AA99}\textbf{4} \\

@@

                                     & \cellcolor[HTML]{99FF99}\textbf{7}

@@

                                     & \cellcolor[HTML]{99FF99}\textbf{17}
